---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loadingData, echo=TRUE}
dataset <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(dataset, "activity.zip", method="curl")
unzip("activity.zip") 
dat <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
```{r meanTotalSteps, echo=TRUE}
# Ignore missing values
dailyData <-  aggregate(steps~date, data=dat, sum, na.rm=TRUE)
dailyMeanStep <- mean(dailyData$steps)
dailyMedianStep <- median(dailyData$steps)
hist(dailyData$steps, main="Total Steps per Day", xlab="Number of Steps", col="gray80")
rug(dailyData$steps, col="gray50")
abline(v= dailyMeanStep, lty=2, col="blue")
abline(v= dailyMedianStep, lty=3, col="red")
text(dailyMeanStep,22,labels=paste("mean:", round(dailyMeanStep)), pos=3, col="blue")
text(dailyMedianStep,24,labels=paste("Median:", round(dailyMedianStep)), pos=3, col="red")
summary(dailyData)
```
From the histogram above, we can tell that:  
The **Mean** total number of steps taken per day is **10766**.  
The **Median** total number of steps taken per day is **10765**.  


## What is the average daily activity pattern?
```{r averageDailyActivity, echo=TRUE}
# Ignore missing values
dailyActivity <- aggregate(steps~interval, data=dat, mean, na.rm=TRUE)
dailyMaxActivity <- max(dailyActivity$steps)
intervalMaxActivity <- dailyActivity[which.max(dailyActivity$steps),]$interval
plot(dailyActivity, type="l", main="Average Steps per day", xlab="Interval Index", ylab="Average Steps taken across the days", col="gray50")
abline(v= intervalMaxActivity, lty=2, col="blue")
abline(h= round(dailyMaxActivity), lty=2, col="blue")
text(intervalMaxActivity, 195,labels=paste("max steps:", round(dailyMaxActivity)), pos=4, col="blue")
text(intervalMaxActivity, 180,labels=paste("interval index:", intervalMaxActivity), pos=4, col="blue")
```

Index: `r intervalMaxActivity` (5-minute) interval, on average across all the days in the dataset, contains the maximum value of `r round(dailyMaxActivity)` number of steps.
 

## Imputing missing values
```{r calculateMissingValues, echo=TRUE}
# Calculate total missing values in the dataset
completeData <- dat[complete.cases(dat),]
incompleteData <- dat[!complete.cases(dat),]
```
Total missing values in dataset: `r nrow(incompleteData)`.  
Total complete values in dataset: `r nrow(completeData)`.


```{r processMissingValues, echo=TRUE}
# Plot 2 charts: original and modified with replaced missing values
par(mfrow=c(1,2))

dailyData <-  aggregate(steps~date, data=dat, sum, na.rm=TRUE)
dailyMeanStep <- mean(dailyData$steps)
dailyMedianStep <- median(dailyData$steps)
hist(dailyData$steps, main="Total Steps per Day", xlab="Number of Steps", ylim=c(0,35), col="gray")
rug(dailyData$steps, col="gray50")
abline(v= dailyMeanStep, lty=2, col="blue")
abline(v= dailyMedianStep, lty=3, col="red")
text(dailyMeanStep,22,labels=paste("mean:",round(dailyMeanStep)), pos=3, col="blue")
text(dailyMedianStep,24,labels=paste("Median:", round(dailyMedianStep)), pos=3, col="red")

# Replace NA values with the average values from the interval
for(i in 1:nrow(dat)){
  if(is.na(dat[i,]$step))
    dat[i,]$steps <- dailyActivity[dailyActivity$interval==(dat[i,]$interval),]$steps
}
newDailyData <-  aggregate(steps~date, data=dat, sum)
newDailyMeanStep <- mean(newDailyData$steps)
newDailyMedianStep <- median(newDailyData$steps)
hist(newDailyData$steps, main="New: Total Steps per Day", xlab="Number of Steps", col="gray")
rug(newDailyData$steps, col="gray50")
abline(v= newDailyMeanStep, lty=2, col="blue")
abline(v= newDailyMedianStep, lty=3, col="red")
text(newDailyMeanStep,22,labels=paste("mean:",round(newDailyMeanStep)), pos=3, col="blue")
text(newDailyMedianStep,24,labels=paste("medum:",round(newDailyMedianStep)), pos=3, col="red")
``` 

The **Mean** total number of steps taken per day is **10766**.  
The **Median** total number of steps taken per day is **10765**.  


```{r summaryData, echo=TRUE}
summary(dailyData) 
summary(newDailyData) 
```
Both the **Mean** and **Median** are the same for both charts.  

The impact of replacing the missing data does not change the **Mean** and **Median** of the datasets.  
The only difference is that it has a **higher 1st Quantile** and a **lower 3rd Quantile**.  


## Are there differences in activity patterns between weekdays and weekends?
```{r fig.height=10, echo=TRUE}
# Panel 2 plot charts
par(mfrow=c(2,1))

# Convert date variable to date type
dat$date <- as.Date(dat$date, format="%Y-%m-%d")
# Add a day and dayType variable
dat$day <- weekdays(dat$date)
dat$dayType <- "weekday"
# Change dayType to weekend if day variable is a Saturday or Sunday.
for(i in 1:nrow(dat)){
  if(dat[i,]$day %in% c("Saturday", "Sunday"))
    dat[i,]$dayType <- "weekend"
}
dat$dayType <- as.factor(dat$dayType)
newActivityData <- aggregate(steps~interval+dayType, dat, mean)
plot(newActivityData[newActivityData$dayType=="weekday",]$steps , type="l", col="brown", main="Weekday", xlab="intervals", ylab="Steps", ylim=c(1,250))
plot(newActivityData[newActivityData$dayType=="weekend",]$steps, type="l", col="brown", main="Weekend", xlab="intervals", ylab="Steps", ylim=c(1,250) )
```